Interfacing logic programming and Coq

### 2020-10-29 papers/types2020/app.tex
some work on sec 3 and 4

Stuff we may not have space for in the paper
\begin{itemize}
\item The LJF sequent rules
\item the code for
  \lsti{dep_pbt}

\item \dots
\end{itemize}
%%% Local Variables:
%%% mode: latex
%%% TeX-master: "main"
%%% End:

>>>>

Stuff we may not have space for in the paper

In Fig.~\ref{fig:interp} we give a proof system for Horn logic and its encoding in coq-elpi
\todo{explain the interface, quotations etc}

\begin{figure}
\newcommand{\XXi}{\Pscr}
\newcommand{\bc}[2]{\Pscr\Downarrow #1 \vdash #2}
\[
\infer{\XXi\vdash t = t}
{%\eqExpert{\XXi}
}
\qquad
\infer{\XXi\vdash \true}
{%\trueExpert{\XXi}
}
\qquad
\infer{\XXi\vdash G_1\vee G_2}
{\XXi\vdash G_i%\qquad \orExpert{\XXi}{\XXi'}{i}
}
\qquad
\infer{\XXi\vdash \exists x. G}
{\XXi\vdash G[t/x]%\qquad \someExpert{\XXi}{\XXi'}{t}
}
\qquad 
\infer{\XXi\vdash G_1\wedge G_2}
{\XXi\vdash G_1\quad \XXi\vdash G_2%\quad \andExpert{\XXi}{\XXi_1}{\XXi_2}
}
\]
\vskip -18pt
\[
\infer{\XXi\vdash A}
      {\bc D A \quad D \in \Pscr}\qquad
  \infer{\bc A A}{}{}\qquad
  \infer{\bc{\forall x.\, D} A}{\bc{D[t/x]} A}{}\qquad
  \infer{\bc{G\supset D} A}{\bc D A\quad \XXi\vdash G}{}
\]
In the judgment $\bc{D}{A}$, the variable $A$ ranges over only atomic
formulas.
\lstinputlisting[linerange={interp-end}]{code/dep-kernel.mod}
\caption{A proof system for Horn logic and its encoding in coq-elpi.}
\label{fig:interp}
%\lstinputlisting[linerange={sigs-end}]{code/dep-kernel.sig}
%\lstinputlisting[linerange={check-end}]{code/dep-kernel.mod}
%\caption{A dependent kernel for Coq terms.}
%\label{fig:kernel}
\end{figure}

%%% Local Variables:
%%% mode: latex
%%% TeX-master: "main"
%%% End:

### 2020-10-29 papers/types2020/pbt.tex
some work on sec 3 and 4

\begin{metanote}
  I'm just saying the bare minimum -am
\end{metanote}

@@

\begin{metanote}
  TODO next
  \begin{itemize}
  \item The elpi dep-pbt tactic
  \item  the simple ordered example [DONE]
  \item a bit of details about the \texttt{aexp} example
  \end{itemize}
\end{metanote}

@@

Whithin the tactic the user
specifies which variables of the environment should be used for
generating data and which for executing the specification. In addition
to this, the user should specify all the certificate information that
will guide the data generation phase. The goal of the current
environment will directly provide the property for which we want to
find a counterexample. % Say we want to generate a counterexample for
% some specification represented by \lsti|H1| and \lsti|H2| in the
% environment, and we want to generate the data for some variables
% \lsti|v1| and \lsti|v2| with a height bounded to 10: the final call to
% the tactic will then look like:
% \begin{lstlisting}
%   elpi pbt 10 (H1 /\ H2) (v1) (v2)
% \end{lstlisting}

To exemplify, let us add to the previous specification a definition of ordered list:
\begin{lstlisting}
Inductive ordered : list nat -> Prop :=
| o_n : ordered []
| o_s : forall x : nat, ordered [x]
| o_c : forall (x y : nat), forall xs,  ordered xs  -> x <= y -> ordered (x:: y :: xs).     
\end{lstlisting}
%
A property we may wish to check before embarking in a formal proof is whether insertion preserves ordered-ness:
\begin{lstlisting}
Conjecture ins_ord:  forall x xs rs, ordered xs -> insert x xs rs -> ordered rs.
intros x xs rs Ho Hi.
elpi dep_pbt height 5 (Ho /\ Hi ) (x) (xs).
Abort.
\end{lstlisting}
In this query the tactic tests the hypotheses \lsti{Ho} and \lsti{Hi}
against data \lsti{x,xs} generated exhaustively up to proofs bounded
by at most $5$ unfoldings of the library \lsti{Inductive} definition
of \lsti{nat} and \lsti{list}. Since we did slip in an error (where?), our
tactic reports a counter-example, namely
\verb|(* Proof Term: [0, [0; 1; 0]] *)|

In order to generate the query to the \lsti|check| predicate, some
pre-processing is needed. In particular, we need to turn the data
variables into \lP logic variables when they appear inside a
specification, and to generate querys to \lsti|check| for each of
these logic variables in association with the type of the data
variable it corresponds to.  In order to code this preprocessing step,
we leverage extensively \lP's higher order programming features. The
substitutions are handled with the technique of \lsti|copy| clauses.

Note that the same query in \textsf{QuickChick} would have required
much more setup: if we wished to proceed relationally as above we
would have had to provide a proof of decidability of the relevant
notions. Were we to use functions, then we would have to implement a
generator and shrinker for ordered lists.
%%% Local Variables:
%%% mode: latex
%%% TeX-master: "main"
%%% End:

>>>>

% \begin{metanote}
%   I'm just saying the bare minimum -am
% \end{metanote}

@@

% \begin{metanote}
%   TODO next
%   \begin{itemize}
%   \item The elpi dep-pbt tactic
%   \item  the simple ordered example [DONE]
%   \item a bit of details about the \texttt{aexp} example
%   \end{itemize}
% \end{metanote}

@@

Whithin the tactic the user
specifies which variables of the environment should be used for
generating data and which for executing the specification. In addition
to this, the user should specify all the certificate information that
will guide the data generation phase. The goal of the current
environment will directly provide the property for which we want to
find a counterexample. % Say we want to generate a counterexample for
% some specification represented by \lsti|H1| and \lsti|H2| in the
% environment, and we want to generate the data for some variables
% \lsti|v1| and \lsti|v2| with a height bounded to 10: the final call to
% the tactic will then look like:
% \begin{lstlisting}
%   elpi pbt 10 (H1 /\ H2) (v1) (v2)
% \end{lstlisting}
Formally, for a property
\[
\forall x_1 \colon \tau_1\dots x_n \colon \tau_m, A_1\land\dots\land A_m \supset B
\]
and the specification of a FPC, the call will be:
\begin{lstlisting}
elpi dep_pbt <fpc> $(A_1\land\dots\land A_m) \ (x_1) \dots (x_n)$.
\end{lstlisting}

To exemplify, let us add to the previous specification of list
insertion a definition of ordered list:
\begin{lstlisting}
Inductive ordered : list nat -> Prop :=
| o_n : ordered []
| o_s : forall x : nat, ordered [x]
| o_c : forall (x y : nat), forall xs,  ordered xs  -> x <= y -> ordered (x:: y :: xs).     
\end{lstlisting}
%
A property we may wish to check before embarking in a formal proof is whether insertion preserves ordered-ness:
\begin{lstlisting}
Conjecture ins_ord:  forall x xs rs, ordered xs -> insert x xs rs -> ordered rs.
intros x xs rs Ho Hi.
elpi dep_pbt height 5 (Ho /\ Hi ) (x) (xs).
Abort.
\end{lstlisting}
In this query the tactic tests the hypotheses \lsti{Ho} and \lsti{Hi}
against data \lsti{x,xs} generated exhaustively up to proofs bounded
by at most $5$ unfoldings of the library \lsti{Inductive} definition
of \lsti{nat} and \lsti{list}. Since we did slip in an error (where?), our
tactic reports a counter-example, namely
\verb|(* Proof Term: [0, [0; 1; 0]] *)|

In order to generate the query to the \lsti|check| predicate, some
pre-processing is needed. In particular, we need to turn the data
variables into \lP logic variables when they appear inside a
specification, and to generate queries to \lsti|check| for each of
these logic variables in association with the type of the data
variable it corresponds to.  In order to realize this preprocessing step,
we leverage extensively \lP's higher order programming features. The
substitutions are handled with the technique of \lsti|copy| clauses.

Note that the same query in \textsf{QuickChick} would have required
much more setup: if we wished to proceed relationally as above, we
would have had to provide a proof of decidability of the relevant
notions. Were we to use functions, then we would have to implement a
generator and shrinker for ordered lists.

\todo{the \texttt{aexp} example}

%%% Local Variables:
%%% mode: latex
%%% TeX-master: "main"
%%% End:

### 2020-10-29 papers/types2020/sec3.tex
some work on sec 3 and 4

\begin{metanote}
  (AM): Start with the logical reconstruction of \texttt{check}, to and then reverse engineer what we want/need to say about focusing and FPC. We \textbf{do} not talk about PBT yet, just logic programming (with proof terms). In sec 4 we package \texttt{check} in a PBT query and explain how \texttt{check} works for term generation.
 One way to proceed:
  \begin{itemize}
  \item One figure fig \ref{fig:interp} for Horn logic over Coq terms (no certificates) and its encoding with quotations 
  \item Then the logic behind the dependent kernel with certificates and proof terms
  \end{itemize}

@@

\subsection{Horn logic in coq-elpi}

@@

In Fig.~\ref{fig:interp} we give a proof system for Horn logic and its encoding in coq-elpi
\todo{explain the interface, quotations etc}

\begin{figure}
\newcommand{\XXi}{\Pscr}
\newcommand{\bc}[2]{\Pscr\Downarrow #1 \vdash #2}
\[
\infer{\XXi\vdash t = t}
{%\eqExpert{\XXi}
}
\qquad
\infer{\XXi\vdash \true}
{%\trueExpert{\XXi}
}
\qquad
\infer{\XXi\vdash G_1\vee G_2}
{\XXi\vdash G_i%\qquad \orExpert{\XXi}{\XXi'}{i}
}
\qquad
\infer{\XXi\vdash \exists x. G}
{\XXi\vdash G[t/x]%\qquad \someExpert{\XXi}{\XXi'}{t}
}
\qquad 
\infer{\XXi\vdash G_1\wedge G_2}
{\XXi\vdash G_1\quad \XXi\vdash G_2%\quad \andExpert{\XXi}{\XXi_1}{\XXi_2}
}
\]
\vskip -18pt
\[
\infer{\XXi\vdash A}
      {\bc D A \quad D \in \Pscr}\qquad
  \infer{\bc A A}{}{}\qquad
  \infer{\bc{\forall x.\, D} A}{\bc{D[t/x]} A}{}\qquad
  \infer{\bc{G\supset D} A}{\bc D A\quad \XXi\vdash G}{}
\]
In the judgment $\bc{D}{A}$, the variable $A$ ranges over only atomic
formulas.
\lstinputlisting[linerange={interp-end}]{code/dep-kernel.mod}
\caption{A proof system for Horn logic and its encoding in coq-elpi.}
\label{fig:interp}
%\lstinputlisting[linerange={sigs-end}]{code/dep-kernel.sig}
%\lstinputlisting[linerange={check-end}]{code/dep-kernel.mod}
%\caption{A dependent kernel for Coq terms.}
%\label{fig:kernel}
\end{figure}

\subsection{Focused proof systems with certificates}
\label{ssec:focused}
\todo{Adapt the discussion}
\todo{Talk about the LJT thing: grammar for formulae and proof terms\cite{Herbelin94,LengrandDM06}?}

@@

\paragraph{Certificates}

\todo{Shorten it}
\todo{Show the dprolog tactic? have an example? could  be
\lsti{Goal  exists R, insert 2 [0;1] R.} and we would move the code for insert here}
{\color{red}
As we mentioned in the introduction, our notion of proof certificates
 is taken from the general setting of \emph{foundational
  proof certificates}~\cite{chihani17jar}.
%
In our case here, an FPC is a collection of \lP clauses that
provide the remaining details not supplied in Figure~\ref{fig:kernel}:
that is, the exact set of constructors for the \lsti{cert} type as
well as the exact specification of the six expert predicates listed in
that figure.
%
Figure~\ref{fig:resources} displays two such FPCs,
both of which can be used to describe proofs where we bound
the number of occurrences of unfoldings in a proof.
%
For example, the first FPC
provides the experts for treating certificates that are constructed
using the \lsti{qheight} constructor.
%
As is easy to verify, the query \mbox{\lsti{(check (qheight 5) B)}} (for the
encoding \lsti{B} of a goal formula) is provable in \lP using the
clauses in Figures~\ref{fig:kernel} and~\ref{fig:resources} if and
only if the height of that proof is 5 or less.
%
Similarly, the second FPC uses the constructor \lsti{qsize} (with two
integers) and can be used to bound the total number of instances of
unfoldings in a proof.
%
In particular, the query \mbox{\lsti{(check (qsize 5 H) B)}}
% \begin{lstlisting}
% sigma H\ (check (qsize 5 H) B)
% \end{lstlisting}
is provable if and only if the total number of unfoldings of that
proof is 5 or less.

@@

\paragraph{A prolog tactic}

@@

\begin{lstlisting}
Elpi Accumulate lp:{{
  solve [str ''height'', int N] [goal _Ctx Ev G _Who] _OutGoals :-
    coq.say "Goal:" {coq.term->string G},
    check (qheight N) (go G Term),
    Ev = Term,
    coq.say "Proof:" {coq.term->string Ev}.
...
}}.
\end{lstlisting}
The following example show how we can do FPC-driven logic programming in Coq and return
a Coq term
\begin{lstlisting}
Inductive insert (x:nat) : list nat -> list nat -> Prop :=
| i_n : is_nat x -> insert x [] [x]
| i_s : forall y: nat, forall ys, x <= y -> insert x (y :: ys) (x :: y :: ys)
| i_c : forall y: nat, forall ys rs, y <= x -> insert x ys rs -> insert x (y :: ys) (x :: rs).
Lemma i1:  exists R, insert 2 [0;1] R.
elpi  dprolog 10.
Qed.
Print i1.
\end{lstlisting}
\begin{verbatim}
ex_intro (fun R : list nat => insert 2 [0; 1] R) [2; 2; 2]
  (i_c 2 0 [1] [2; 2] (le_S 0 1 (le_S 0 0 (le_n 0)))
	 (i_c 2 1 [] [2] (le_S 1 1 (le_n 1)) (i_n 2)))
     : exists R : list nat, insert 2 [0; 1] R
\end{verbatim}
%
%%% Local Variables:
%%% mode: latex
%%% TeX-master: "main"
%%% End:

>>>>

\begin{metanote}
  (AM): Start with the logical reconstruction of \texttt{check}, to and then reverse engineer what we want/need to say about focusing and FPC. We \textbf{do} not talk about PBT yet, just logic programming (with proof terms). In sec 4 we package \texttt{check} in a PBT query and explain how \texttt{check} works for term generation.
 One way to proceed:
  % \begin{itemize}
  % \item One figure fig \ref{fig:interp} for Horn logic over Coq terms (no certificates) and its encoding with quotations 
  % \item Then the logic behind the dependent kernel with certificates and proof terms
  % \end{itemize}

@@

%\subsection{Horn logic in coq-elpi}

\subsection{Focused proof systems with certificates}
\label{ssec:focused}
\todo{Adapt the discussion}
\todo{grammar for formulae and proof terms\cite{Herbelin94,LengrandDM06}?}

@@

@@

\subsection{Certificates}

\todo{Shorten it}
% \todo{Show the dprolog tactic? have an example? could  be
% \lsti{Goal  exists R, insert 2 [0;1] R.} and we would move the code for insert here}
{\color{red}
As we mentioned in the introduction, our notion of proof certificates
 is taken from the general setting of \emph{foundational
  proof certificates}~\cite{chihani17jar}.
%
In our case here, an FPC is a collection of \lP clauses that
provide the remaining details not supplied in Figure~\ref{fig:kernel}:
that is, the exact set of constructors for the \lsti{cert} type as
well as the exact specification of the six expert predicates listed in
that figure.
%
Figure~\ref{fig:resources} displays two such FPCs,
both of which can be used to describe proofs where we bound
the number of occurrences of unfoldings in a proof.
%
For example, the first FPC
provides the experts for treating certificates that are constructed
using the \lsti{qheight} constructor.
%
As is easy to verify, the query \mbox{\lsti{(check (qheight 5) B)}} (for the
encoding \lsti{B} of a goal formula) is provable in \lP using the
clauses in Figures~\ref{fig:kernel} and~\ref{fig:resources} if and
only if the height of that proof is 5 or less.
%
Similarly, the second FPC uses the constructor \lsti{qsize} (with two
integers) and can be used to bound the total number of instances of
unfoldings in a proof.
%
In particular, the query \mbox{\lsti{(check (qsize 5 H) B)}}
% \begin{lstlisting}
% sigma H\ (check (qsize 5 H) B)
% \end{lstlisting}
is provable if and only if the total number of unfoldings of that
proof is 5 or less.

@@

\subsection{A prolog tactic}

@@

\begin{lstlisting}
Elpi Accumulate lp:{{
  solve [str ''height'', int N] [goal _Ctx Ev G _Who] _OutGoals :-
    coq.say "Goal:" {coq.term->string G},
    check (qheight N) (go G Term),
    Ev = Term,
    coq.say "Proof:" {coq.term->string Ev}.
...
}}.
\end{lstlisting}
The following example show how we can do FPC-driven logic programming in Coq and return
a Coq term
\begin{lstlisting}
Inductive insert (x:nat) : list nat -> list nat -> Prop :=
| i_n : is_nat x -> insert x [] [x]
| i_s : forall y: nat, forall ys, x <= y -> insert x (y :: ys) (x :: y :: ys)
| i_c : forall y: nat, forall ys rs, y <= x -> insert x ys rs -> insert x (y :: ys) (x :: rs).
Lemma i1:  exists R, insert 2 [0;1] R.
eexists.
elpi  dprolog 10.
Qed.
Print i1.
\end{lstlisting}
\begin{verbatim}
ex_intro (fun R : list nat => insert 2 [0; 1] R) [2; 2; 2]
  (i_c 2 0 [1] [2; 2] (le_S 0 1 (le_S 0 0 (le_n 0)))
	 (i_c 2 1 [] [2] (le_S 1 1 (le_n 1)) (i_n 2)))
     : exists R : list nat, insert 2 [0; 1] R
\end{verbatim}
%
%%% Local Variables:
%%% mode: latex
%%% TeX-master: "main"
%%% End:
